---
title: "Class09: Halloween Candy Mini Project"
author: "Lilith Sadil, A16470107"
format: html
--- 

### Class 9 - 4.30.24

# Halloween Candy Mini Project

## 1. Importing Candy Data

In today's class, we'll examinne data about candy from the "538" website. We'll start by importing the data:

```{r}
candy_file <- "candy-data.txt"
```

Now, we'll turn the

```{r}
candy = read.csv(candy_file, row.names=1)
head(candy)
```

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

There are 85 different candies in the data set.

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```
There are 38 fruity candies in the data set.

## 2. What's Your Favorite Candy?

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

My favorite candy mentioned in the dataset is the Reese's Peanut Butter cup.
```{r}
candy["Reese's Peanut Butter cup", ]$winpercent
```
The win percent of the Reese's Peanut Butter cup is 84.18%.

>> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```
The win percent for Kit Kat is 76.77%.

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```
The winpercent of Tootsie Roll Snack Bars is 49.65%.

There's a package called "skimr" which can help give a quick overview of a datset
```{r}
#install.packages("skimr")
library("skimr")
skimr::skim(candy)
```
> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

It appears that winpercent is measured on a different scale compared to the the rest of the rows.

> Q7. What do you think a zero and one represent for the candy$chocolate column?

"0" and "1" in the candy$chocolate column likely refer to whether or not a candy contains chocolate (0 for "no" and 1 for "yes").

> Q8. Plot a histogram of winpercent values

To make a simple plot, we can use the `hist` operation in R:
```{r}
hist(candy$winpercent)
```

We can recreate this plot using ggplot:
```{r}
library(ggplot2)

win_hist = ggplot(candy) +
  aes(candy$winpercent)+
  geom_histogram()
win_hist
```

> Q9. Is the distribution of winpercent values symmetrical?

The distribution of win percent values is not symmetrical; it's slightly skewed towards the left with a tail on the right.

> Q10. Is the center of the distribution above or below 50%?

```{r}
summary(candy$winpercent)
```
The center of the distribution (median) lies at 47.83% which is roughly in the middle.

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

1. We have to find all the chocolate candies,
2. Find their winpercents,
3. And compare the mean of their winpercents
```{r}
chocolate_candy_win = candy$winpercent[as.logical(candy$chocolate)]
# or chocolate_candy = candy$winpercent[candy$chocolate==1]
chocolate_candy_win
```
This has provided us with the win percents of only the candies containing chocolate. Now we can take the mean and compare them to fruit candies

```{r}
mean.choc = mean(chocolate_candy_win)
mean.choc
```
The average winpercent of candy containing chocolate is 60.92%
```{r}
fruit_candy_win = candy$winpercent[as.logical(candy$fruity)]
mean.fruit = mean(fruit_candy_win)
mean.fruit
```
The average winpercent of candy containing fruit is 44.12%. From this, we know that chocolate candies, on average, tend to be preferred over fruity candies.

> Q12. Is this difference statistically significant?

```{r}
t.test(mean.choc, mean.fruit)
```

## 3. Overall Candy Rankings
