---
title: "Class 18 - Pertussis and the CMI-PB project"
author: "Lilith Sadil"
format: gfm
---

## Background

Pertussis (whooping cough) is a highly contagious lung infection which is most deadly among infants.

## Investigating pertussis cases by year

Let's start by looking at the number of Pertussis cases/year in the U.S.

The CDC tracks Pertussis cases, making the data available here:

https://www.cdc.gov/pertussis/php/surveillance/pertussis-cases-by-year.html

We can use a package called `datapasta` to extract tables/data from a web page where a file for the data is not provided:

```{r}
library(datapasta)
cdc = data.frame(year = c(1922L,1923L,1924L,1925L,1926L,
                                         1927L,1928L,1929L,1930L,1931L,
                                         1932L,1933L,1934L,1935L,1936L,1937L,
                                         1938L,1939L,1940L,1941L,1942L,
                                         1943L,1944L,1945L,1946L,1947L,1948L,
                                         1949L,1950L,1951L,1952L,1953L,1954L,
                                         1955L,1956L,1957L,1958L,1959L,
                                         1960L,1961L,1962L,1963L,1964L,1965L,
                                         1966L,1967L,1968L,1969L,1970L,
                                         1971L,1972L,1973L,1974L,1975L,1976L,
                                         1977L,1978L,1979L,1980L,1981L,
                                         1982L,1983L,1984L,1985L,1986L,1987L,
                                         1988L,1989L,1990L,1991L,1992L,1993L,
                                         1994L,1995L,1996L,1997L,1998L,
                                         1999L,2000L,2001L,2002L,2003L,2004L,
                                         2005L,2006L,2007L,2008L,2009L,
                                         2010L,2011L,2012L,2013L,2014L,2015L,
                                         2016L,2017L,2018L,2019L,2020L,
                                         2021L),
        cases = c(107473,164191,165418,152003,
                                         202210,181411,161799,197371,166914,
                                         172559,215343,179135,265269,180518,
                                         147237,214652,227319,103188,183866,
                                         222202,191383,191890,109873,133792,
                                         109860,156517,74715,69479,120718,
                                         68687,45030,37129,60886,62786,
                                         31732,28295,32148,40005,14809,11468,
                                         17749,17135,13005,6799,7717,9718,
                                         4810,3285,4249,3036,3287,1759,
                                         2402,1738,1010,2177,2063,1623,1730,
                                         1248,1895,2463,2276,3589,4195,
                                         2823,3450,4157,4570,2719,4083,6586,
                                         4617,5137,7796,6564,7405,7298,
                                         7867,7580,9771,11647,25827,25616,
                                         15632,10454,13278,16858,27550,18719,
                                         48277,28639,32971,20762,17972,
                                         18975,15609,18617,6124,2116)
      )
```

## A tale of two vaccines (wP & aP)

> Q2. Now that we've added the CDC table in as a data frame, let's make a graph of the cases per year:

```{r}
library(ggplot2)
ggplot(cdc) +
  aes(x=year,y=cases) + 
  geom_line(col = "grey")+
  geom_point(col="black")+
  geom_vline(xintercept=1946, col="blue", linetype=2)+
  geom_vline(xintercept = 1996, col="red", linetype=2)+
  geom_vline(xintercept = 2019, col="orange", linetype=2)
```

Two vertical lines have been included to show where the wP & aP vaccines were implemented.

> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

Shortly after the aP vaccine was introduced in \~2000, the anti-vaccine movement began which could contribute to the rise again in cases around the time. Additionally, the acellular vaccine could have slightly lower efficacy since it doesn't cause the same strength of antigen response as the the full cell vaccine.

## Exploring CMI-PB data

CMI-PB is a systems vaccinology project which focuses on discerning the differences between aP and wP immune responses:

The resource has an API (application programming interface) which returns JSON formatted data in key-value pairs:

```{r}
library(jsonlite)
subject = read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 
head(subject, 3) #view information from the first three subjects
```

> How many individuals are in the dataset?

```{r}
nrow(subject)
```

There are 118 individuals in the dataset.

> Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```

60 individuals were infants vaccinated with the aP vaccine & 58 individuals were infants vaccinated with the wP vaccine

> Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```

There were 79 female participants and 39 male participants in the study.

> Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```

The table generated above displays the breakdown of race & gender in the study.

Next, we'll download some more data from the CMI-PB site:

```{r}
specimen = read_json("http://cmi-pb.org/api/specimen", simplifyVector = TRUE) 
ab_titer = read_json("http://cmi-pb.org/api/v4/plasma_ab_titer", simplifyVector=TRUE)

head(subject, 3)
head(ab_titer, 3)
```

We will use functions from the Dplyr package to join some of these tables together into more comprehensive data sets:

```{r}
library(dplyr)
meta = inner_join(subject, specimen)
head(meta, 3)
```

Now we can join the meta table & our antibody titer table:

```{r}
abdata = inner_join(ab_titer, meta)
head(abdata,3)
```

```{r}
dim(abdata)
```

What antibodies are measured in the `ab_titer` table?

```{r}
unique(ab_titer$isotype)
```

IgE, IgG, IgG1, IgG2, IgG3, and IgG4 are the 6 different isotypes measured in the ab_titer table. The number of each isotype instance is as follows:

```{r}
table(ab_titer$isotype)
```

The different antigens measured in the study are as follows:

```{r}
table(ab_titer$antigen)
```

## Examine IgG Ab Titer Levels

```{r}
igg = abdata %>% filter(isotype == "IgG")
head(igg)
```
```{r}
library(ggplot2)
base_plot = ggplot(igg) +
  aes(MFI, antigen) +
  geom_boxplot() 
base_plot
```
```{r}
base_plot + 
  xlim(0,75) +
  facet_wrap(vars(visit), nrow=2)
```
```{r}
norm_plot = ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot()+  
  xlim(0,75) +
  facet_wrap(vars(visit), nrow=2)
norm_plot
```

```{r}
abdata.21 <- abdata %>% filter(dataset == "2021_dataset")

abdata.21 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```












